---
title: "glmOJ: Count Regression Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{glmOJ: Count Regression Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(glmOJ)
```

## Overview

`glmOJ` provides a streamlined workflow for fitting, diagnosing, and
interpreting count regression models. The six supported families are:

| Function | Model |
|---|---|
| `poissonGLM()` | Poisson GLM |
| `negbinGLM()` | Negative Binomial GLM |
| `tweedieGLM()` | Tweedie GLM |
| `zeroinflPoissonGLM()` | Zero-Inflated Poisson |
| `zeroinflNegbinGLM()` | Zero-Inflated Negative Binomial |
| `zeroinflTweedieGLM()` | Zero-Inflated Tweedie |

A general-purpose wrapper `countGLM()` fits all six and selects the best by
a user-chosen criterion (`decide`): `"BIC"` (default), `"AIC"`, `"LogLik"`,
or `"McFadden"` (McFadden pseudo-R²). Each zero-inflated counterpart
(Poisson, Negative Binomial, Tweedie) is fitted only when the DHARMa
zero-inflation test flags its base model (p < 0.05).

---

## Case Study: Federal Environmental Crime Prosecutions

Greenberg et al. (2026) investigate how environmental and social factors
influence where EPA criminal prosecutions occur across 3,143 US counties
(2011–2020). The response variable `FinalEC` is a count of criminal
prosecutions per county.

```{r load-data}
data("Greenberg26.dat")
```

### 1. Data Exploration

Before fitting, `summarizeCountData()` gives a quick numerical and graphical
overview of the count response alongside each predictor.

```{r summarize, fig.height = 6, cache = TRUE}
summarizeCountData(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
```

### 2. Poisson Regression

We first fit a Poisson GLM with the Overall Environmental Quality Index and
demographic/geographic controls.

```{r poisson}
mod.pois <- poissonGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
```

#### Coefficients (exponentiated)

```{r pois-coef}
mod.pois$coefficients
```

#### Model fit

```{r pois-diag, fig.height = 4}
mod.pois$diagnostics$plot
```

The dispersion ratio of `r round(mod.pois$diagnostics$dispersion_ratio, 3)`
is flagged in red — the observed variance is ~60% larger than expected under
Poisson, suggesting overdispersion. We also inspect the squared Pearson
residual plot:

```{r pois-r2, fig.height = 4, fig.width = 5}
mod.pois$diagnostics$r2_plot
```

The wedge shape confirms the mean-variance relationship is not well captured
by the Poisson assumption.

### 3. Negative Binomial Regression

The negative binomial adds a free dispersion parameter $\theta$ to handle
overdispersion.

```{r negbin}
mod.nb <- negbinGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat,
  control = stats::glm.control(maxit = 100)
)
```

#### Coefficients (exponentiated)

```{r nb-coef}
mod.nb$coefficients
```

#### Model fit

```{r nb-diag, fig.height = 4}
mod.nb$diagnostics$plot
```

```{r nb-r2, fig.height = 4, fig.width = 5}
mod.nb$diagnostics$r2_plot
```

The dispersion ratio is now
`r round(mod.nb$diagnostics$dispersion_ratio, 3)` — much closer to 1. The
estimated $\theta$ is `r round(mod.nb$theta, 3)`.

### 4. Using the `countGLM` Wrapper

Rather than fitting each model manually, `countGLM()` fits all six families
at once and selects the best by the criterion specified in `decide` (default
`"BIC"`) — arriving at the same conclusion automatically.

#### Overall EQI formula

```{r countglm1}
result1 <- countGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
print(result1)
```

The wrapper selects the same winner as the manual LRT. Individual fits remain
accessible: `result1$fits$negbin`, `result1$fits$poisson`, etc.

#### Sub-index formula

The researchers also evaluated whether separate Water, Air, Land, and
Socioeconomic indices were more informative than the composite Overall EQI.
`countGLM` handles this in one call:

```{r countglm2}
result2 <- countGLM(
  FinalEC ~ water_eqi_2jan2018_vc +
    land_eqi_2jan2018_vc +
    air_eqi_2jan2018_vc +
    sociod_eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
print(result2)
```

Again the negative binomial is selected. The non-nested comparison between
these two winning models (overall EQI vs sub-indices) can be done with a Vuong
test via `nonnest2::vuongtest(result1$fits$negbin$model, result2$fits$negbin$model)`.

---

## 5. Coefficient Interpretation

`interpret_coef()` translates any exponentiated coefficient into a
plain-language statement, with a 95% CI and an automatic note when the
predictor is not discernible from zero (p > 0.05).

#### Significant predictor

```{r interp-pct}
interpret_coef(mod.nb, "pctnonwhite10")
```

#### Non-significant predictor

```{r interp-eqi}
interpret_coef(mod.nb, "eqi_2jan2018_vc")
```

The note about discernibility is added automatically.

#### Using with `countGLM`

Pass the `countGLM` result directly — it delegates to the best-fitting model:

```{r interp-countglm}
interpret_coef(result2, "pctnonwhite10")
```

#### Zero-inflated models: specifying the component

For zero-inflated models, use `component = "count"` (default) or
`component = "zero"`:

```{r interp-zi, eval = FALSE}
interpret_coef(zi_fit, "pctnonwhite10", component = "count")
interpret_coef(zi_fit, "pctnonwhite10", component = "zero")
```

---

## Case Study 2: Urban Surveillance Camera Counts (Dahir 2025)

Dahir (2025) models the number of surveillance cameras per census tract
across ten US cities. The response `cam_count` is a non-negative integer
with 87.6% zeros and a variance-to-mean ratio of approximately 3.6, a warning sign that a standard Poisson model may fit poorly.

```{r load-dahir}
data("Dahir25.dat")
```

### 6. Data Exploration

```{r dahir-summarize, fig.height = 5, cache = TRUE}
summarizeCountData(
  cam_count ~ total_crime_rate + hinc + modal_zone + pvac,
  data = Dahir25.dat
)
```

Camera placement varies substantially by land-use zone — mixed and industrial tracts have
cameras in roughly 28–42% of cases, while residential tracts (the
majority, n = 8913) have cameras in fewer than 10%. This mixture of
structurally camera-free tracts alongside genuinely high-count commercial
zones is exactly the setting where zero-inflated or overdispersed count
models are needed.

### 7. Poisson and Zero-Inflated Poisson

We may want to start by manually fitting a Poisson model to confirm the overdispersion and zero-inflation intuition. The formula includes a quadratic term for each predictor to allow for non-linear effects, and road length is included as an offset:

```{r dahir-poisson, warning=TRUE}
pois.cam <- poissonGLM(
  cam_count ~ pnhwht +
    pnhblk +
    entropy_rank +
    total_crime_rate +
    modal_zone +
    pop +
    hinc +
    pvac +
    mhmval +
    city +
    offset(log_road_length) +
    I(pnhwht^2) +
    I(pnhblk^2) +
    I(entropy_rank^2),
  data = Dahir25.dat
)
print(pois.cam)
```

The DHARMa zero-inflation test result:

```{r dahir-poisson-zi}
zi <- pois.cam$diagnostics$zi_test
cat(sprintf(
  "DHARMa zero-inflation test: p = %.4f  |  Detected: %s\n",
  zi$p_value,
  zi$detected
))
zi$plot
```

### 8. Automatic Model Selection with `countGLM`

`countGLM()` fits all six count families and selects the best by the
criterion in `decide` (default `"BIC"`). Road length (`log_road_length`) is
included as an offset in the count component; because `ziformula = NULL`, it
is automatically applied to the zero-inflation component as well — `countGLM`
prints a note confirming this:

```{r dahir-countglm}
result_cam <- countGLM(
  cam_count ~ pnhblk +
    pnhwht +
    total_crime_rate +
    hinc +
    pvac +
    modal_zone +
    offset(log_road_length),
  data = Dahir25.dat
)
print(result_cam)
```

The comparison table shows AIC and BIC for all six families, sorted by the
selection criterion. The selected model is **`r result_cam$best_model`**. The
recommendation captures the relevant dispersion and zero-inflation diagnostics
automatically, explaining why this family was preferred.

### 9. Checking for Multicollinearity (VIF)

All four model families share the assumption that predictors are not strongly
collinear. `countGLM()` automatically computes Variance Inflation Factors and
stores them in `$vif`. To avoid false positives from *structural* collinearity
— which is expected whenever interaction or polynomial terms are present —
VIF is computed on an OLS model that retains only the order-1 (main-effect)
terms from the formula. Offset terms are excluded as well.

```{r dahir-vif}
result_cam$vif
```

VIF = 1 means a predictor is uncorrelated with all others; values up to ~5
are generally acceptable. Any VIF > 5 triggers a warning. Here all
predictors are well below that threshold, confirming that multicollinearity
is not a concern for this model.

To illustrate the interaction-term handling, fitting the Poisson model with
a quadratic term does **not** inflate the VIF for the underlying predictors:

```{r dahir-vif-interaction}
result_quad <- suppressWarnings(countGLM(
  cam_count ~ pnhblk +
    pnhwht +
    total_crime_rate +
    I(total_crime_rate^2) +
    offset(log_road_length),
  data = Dahir25.dat
))
result_quad$vif # only main-effect terms: pnhblk, pnhwht, total_crime_rate
```

### 10. Interpreting the Winning Model

`interpret_coef()` delegates to the best-fitting model automatically:

```{r dahir-interp-crime}
interpret_coef(result_cam, "total_crime_rate")
```

```{r dahir-interp-income}
interpret_coef(result_cam, "hinc")
```

For zero-inflated models, the count and zero components can be interpreted
separately. The count component describes the expected camera count among
tracts that are in the counting process; the zero component describes the
odds of being a structural zero (a tract that never receives a camera):

```{r dahir-interp-zi, eval = inherits(result_cam$fits[[result_cam$best_model]], "zeroinflGLMfit")}
interpret_coef(result_cam, "total_crime_rate", component = "count")
interpret_coef(result_cam, "total_crime_rate", component = "zero")
```

---

## Case Study 3: Zero-Inflated Tweedie (Simulated Count Data)

The `ZITweedie.dat` dataset illustrates when `zeroinflTweedieGLM()` is the
appropriate model. The response `y` is a non-negative integer count simulated
from a compound Poisson-Gamma (Tweedie, $p = 1.5$, $\phi = 2.0$) with
structural zeros added on top. The true model has two independent predictors:
`x1` drives the count mean and `x2` drives zero-inflation.

```{r load-zitw}
data("ZITweedie.dat")
```

### 11. Data Exploration

```{r zitw-summarize, fig.height = 4, cache = TRUE}
summarizeCountData(y ~ x1 + x2, data = ZITweedie.dat)
```

The frequency table is dominated by zeros (~57%), and the variance far exceeds
the mean — both signs that a standard Poisson or negative binomial model will
fit poorly.

### 12. Automatic Model Selection with `countGLM`

`countGLM()` fits the three base families and then fits each zero-inflated
counterpart only when the DHARMa zero-inflation test flags its base model
(p < 0.05). AIC/BIC then arbitrates among whichever models were fit.

```{r zitw-countglm}
result_zitw <- suppressWarnings(
  countGLM(y ~ x1 + x2, data = ZITweedie.dat)
)
print(result_zitw)
```

### 13. Fitting Zero-Inflated Tweedie Directly

`zeroinflTweedieGLM()` can also be called directly, which is useful when the
ZI predictor differs from the count predictor. Here the true ZI predictor is
`x2`, so we supply it via `ziformula`:

```{r zitw-direct}
fit_zitw <- suppressWarnings(
  zeroinflTweedieGLM(y ~ x1, data = ZITweedie.dat, ziformula = ~x2)
)
print(fit_zitw)
```

### 14. Interpreting Both Components

The count component describes the expected count among observations that are
not structural zeros; the zero component describes the odds of being a
structural zero.

```{r zitw-interp-count}
interpret_coef(fit_zitw, "x1", component = "count")
```

```{r zitw-interp-zero}
interpret_coef(fit_zitw, "x2", component = "zero")
```

The estimated $\hat{p}$ (`r round(fit_zitw$p, 3)`) is well within (1, 2),
confirming the Tweedie family is appropriate. The estimated $\hat{\phi}$ is
`r round(fit_zitw$phi, 3)`.

---

## Untangling Interactions

A single coefficient is not enough to describe an interaction: the effect of
one predictor depends on the value of another. `untangle_interaction()`
produces the right post-hoc summary for each case — categorical × categorical,
categorical × continuous, or continuous × continuous — in one call. It
accepts any model fit returned by the package (or a raw `glm` / `glmmTMB`
object) and a pair of interacting variables.

By default the function does **not** average over any *categorical* predictor
outside the interaction of interest; each level is reported separately.
Continuous controls are held at their mean (the `emmeans` default). Pass
names to `average.over` to collapse specific categorical predictors.

We return to the Greenberg environmental-crime data and fit three variants
of the Poisson model, each with a different interaction, to illustrate the
three cases.

### 15. Categorical × Categorical: `metro * EPAregion`

Does the metro/non-metro contrast look the same across EPA regions?

```{r untangle-cat-cat}
fit_cc <- poissonGLM(
  FinalEC ~ metro * EPAregion + eqi_2jan2018_vc + pctnonwhite10 + gdp2017b,
  data = Greenberg26.dat
)

res_cc <- untangle_interaction(fit_cc, c("metro", "EPAregion"))
head(res_cc$emmeans, 10)
```

`Mean` is the estimated marginal mean on the response scale (expected count
per county); `Lower`/`Upper` are 95% confidence bounds. Comparing `Mean`
across `metro` within each `EPAregion` shows how the metro effect varies by
region. The companion plot shows every cell-mean with its 95% error bar
(and is faceted by any categorical predictor left unaveraged — here there
are none, so a single panel is drawn):

```{r untangle-cat-cat-plot, fig.height = 5}
res_cc$plot
```

### 16. Categorical × Continuous: `metro * pctnonwhite10`

How does the effect of percent-non-white-population differ between metro and
non-metro counties? For this case `untangle_interaction()` returns two
pieces:

1.  `emtrends` — the slope of the continuous predictor at each level of the
    factor (on the linear-predictor / link scale).
2.  `plot` — a faceted `ggplot` of predicted responses across 100 evenly
    spaced values of the continuous predictor, coloured by the factor.
    Facets are combinations of any categorical predictor *not* in
    `average.over` and *not* in the interaction.

First, leave every other categorical predictor in the model unaveraged.
`EPAregion` has 10 levels, which would exceed the 8-facet limit, so the
plot is suppressed with a warning:

```{r untangle-cat-cont-warn}
fit_xc <- poissonGLM(
  FinalEC ~ metro * pctnonwhite10 + eqi_2jan2018_vc + gdp2017b + EPAregion,
  data = Greenberg26.dat
)

res_xc_warn <- untangle_interaction(fit_xc, c("metro", "pctnonwhite10"))
res_xc_warn$emtrends
is.null(res_xc_warn$plot)
```

Averaging over `EPAregion` removes the facets and the plot is built
normally:

```{r untangle-cat-cont, fig.height = 4}
res_xc <- untangle_interaction(
  fit_xc,
  c("metro", "pctnonwhite10"),
  average.over = "EPAregion"
)
res_xc$emtrends
res_xc$plot
```

If we want every region retained despite exceeding the facet limit, we can
set `overridePlot = TRUE`:

```{r untangle-cat-cont-override, eval = FALSE}
untangle_interaction(
  fit_xc,
  c("metro", "pctnonwhite10"),
  overridePlot = TRUE
)
```

### 17. Continuous × Continuous: `pctnonwhite10 * gdp2017b`

For two continuous predictors `untangle_interaction()` returns:

1.  Two `emtrends_*` data frames — the slope of each predictor at low
    (mean − SD), medium (mean), and high (mean + SD) values of the other.
2.  Two `johnson_neyman_*` objects — the range of the moderator over which
    the focal slope is statistically discernible, computed via
    `interactions::johnson_neyman()` with FDR control.

```{r untangle-cont-cont, fig.height = 4}
fit_nn <- poissonGLM(
  FinalEC ~ pctnonwhite10 * gdp2017b + metro + EPAregion,
  data = Greenberg26.dat
)

res_nn <- untangle_interaction(fit_nn, c("pctnonwhite10", "gdp2017b"))
head(res_nn$emtrends_pctnonwhite10)
head(res_nn$emtrends_gdp2017b)
```

Johnson-Neyman intervals describe the range of the moderator over which the
focal slope is statistically discernible. Printing an object gives the
numerical summary and renders the shaded-region plot; each object also
carries a `$plot` element you can extract separately:

```{r untangle-jn, fig.height = 4}
res_nn$johnson_neyman_pctnonwhite10_by_gdp2017b
```

The interval shows that the slope of `pctnonwhite10` is significant (p < .05)
when `gdp2017b` is **outside** roughly [104, 261] — meaning the race–prosecution
relationship is detectable in both low- and high-GDP counties, but washes out
at intermediate values. The reversed call (focal = `gdp2017b`) is in
`res_nn$johnson_neyman_gdp2017b_by_pctnonwhite10`.

If the continuous predictors have already been standardized (mean 0, SD 1),
pass `standardized = TRUE` so the low/medium/high grid uses `-1, 0, 1` and
the interpretation text refers to a "one-standard-deviation increase"
rather than a "one-unit increase".

---

## References

Dahir, Abdi Latif. 2025. "Surveillance Camera Placement and Urban
Inequality." Working paper.

Greenberg, Pierce, Erik W Johnson, Jennifer Schwartz, and Rylie Wartinger.
2026. "Social Factors Shape Federal Environmental Crime Prosecution Patterns
in the USA." *Nature Sustainability*, 1–5.
