---
title: "glmOJ: Count Regression Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{glmOJ: Count Regression Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(glmOJ)
```

## Overview

`glmOJ` provides a streamlined workflow for fitting, diagnosing, and
interpreting count regression models. The four supported families are:

| Function | Model |
|---|---|
| `poissonGLM()` | Poisson GLM |
| `negbinGLM()` | Negative Binomial GLM |
| `zeroinflPoissonGLM()` | Zero-Inflated Poisson |
| `zeroinflNegbinGLM()` | Zero-Inflated Negative Binomial |

A general-purpose wrapper `countGLM()` fits all four and selects the best by
AIC.

---

## Case Study: Federal Environmental Crime Prosecutions

Greenberg et al. (2026) investigate how environmental and social factors
influence where EPA criminal prosecutions occur across 3,143 US counties
(2011–2020). The response variable `FinalEC` is a count of criminal
prosecutions per county.

```{r load-data}
data("Greenberg26.dat")
```

### 1. Data Exploration

Before fitting, `summarizeCountData()` gives a quick numerical and graphical
overview of the count response alongside each predictor.

```{r summarize, fig.height = 6}
summarizeCountData(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
```

### 2. Poisson Regression

We first fit a Poisson GLM with the Overall Environmental Quality Index and
demographic/geographic controls.

```{r poisson, cache = TRUE}
mod.pois <- poissonGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
```

#### Coefficients (exponentiated)

```{r pois-coef}
mod.pois$coefficients
```

#### Model fit

```{r pois-diag, fig.height = 4}
mod.pois$diagnostics$plot
```

The dispersion ratio of `r round(mod.pois$diagnostics$dispersion_ratio, 3)`
is flagged in red — the observed variance is ~60% larger than expected under
Poisson, suggesting overdispersion. We also inspect the squared Pearson
residual plot:

```{r pois-r2, fig.height = 4, fig.width = 5}
mod.pois$diagnostics$r2_plot
```

The wedge shape confirms the mean-variance relationship is not well captured
by the Poisson assumption.

### 3. Negative Binomial Regression (Model 1b)

The negative binomial adds a free dispersion parameter $\theta$ to handle
overdispersion.

```{r negbin, cache = TRUE}
mod.nb <- negbinGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat,
  control = stats::glm.control(maxit = 100)
)
```

#### Coefficients (exponentiated)

```{r nb-coef}
mod.nb$coefficients
```

#### Model fit

```{r nb-diag, fig.height = 4}
mod.nb$diagnostics$plot
```

```{r nb-r2, fig.height = 4, fig.width = 5}
mod.nb$diagnostics$r2_plot
```

The dispersion ratio is now
`r round(mod.nb$diagnostics$dispersion_ratio, 3)` — much closer to 1. The
estimated $\theta$ is `r round(mod.nb$theta, 3)`.

### 4. Comparing Models: Likelihood Ratio Test

Because the Poisson model is nested within the negative binomial (Poisson is
NB with $\theta \to \infty$), we can use a likelihood ratio test. The
underlying `glm`/`glm.nb` fit objects are available via `$model`:

```{r lrt}
lmtest::lrtest(mod.pois$model, mod.nb$model)
```

The negative binomial model is significantly better ($p < 0.0001$), confirming
that overdispersion is a genuine problem for the Poisson fit.

### 5. Poisson Regression with Specific Quality Indices 

The researchers also evaluated whether separate Water, Air, Land, and
Socioeconomic indices were more informative than the composite Overall EQI.

```{r pois2, cache = TRUE}
mod.pois2 <- poissonGLM(
  FinalEC ~ water_eqi_2jan2018_vc +
    land_eqi_2jan2018_vc +
    air_eqi_2jan2018_vc +
    sociod_eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)

mod.pois2$diagnostics$plot
```

Dispersion ratio: `r round(mod.pois2$diagnostics$dispersion_ratio, 3)` —
again flagging overdispersion.

### 6. Negative Binomial with Specific Quality Indices

```{r nb2, cache = TRUE}
mod.nb2 <- negbinGLM(
  FinalEC ~ water_eqi_2jan2018_vc +
    land_eqi_2jan2018_vc +
    air_eqi_2jan2018_vc +
    sociod_eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat,
  control = stats::glm.control(maxit = 100)
)

mod.nb2$diagnostics$plot
```

Dispersion ratio: `r round(mod.nb2$diagnostics$dispersion_ratio, 3)`.


---

## References

Greenberg, Pierce, Erik W Johnson, Jennifer Schwartz, and Rylie Wartinger.
2026. "Social Factors Shape Federal Environmental Crime Prosecution Patterns
in the USA." *Nature Sustainability*, 1–5.
