---
title: "glmOJ: Count Regression Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{glmOJ: Count Regression Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(glmOJ)
```

## Overview

`glmOJ` provides a streamlined workflow for fitting, diagnosing, and
interpreting count regression models. The four supported families are:

| Function | Model |
|---|---|
| `poissonGLM()` | Poisson GLM |
| `negbinGLM()` | Negative Binomial GLM |
| `zeroinflPoissonGLM()` | Zero-Inflated Poisson |
| `zeroinflNegbinGLM()` | Zero-Inflated Negative Binomial |

A general-purpose wrapper `countGLM()` fits all four and selects the best by
AIC.

---

## Case Study: Federal Environmental Crime Prosecutions

Greenberg et al. (2026) investigate how environmental and social factors
influence where EPA criminal prosecutions occur across 3,143 US counties
(2011–2020). The response variable `FinalEC` is a count of criminal
prosecutions per county.

```{r load-data}
data("Greenberg26.dat")
```

### 1. Data Exploration

Before fitting, `summarizeCountData()` gives a quick numerical and graphical
overview of the count response alongside each predictor.

```{r summarize, fig.height = 6}
summarizeCountData(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
```

### 2. Poisson Regression

We first fit a Poisson GLM with the Overall Environmental Quality Index and
demographic/geographic controls.

```{r poisson}
mod.pois <- poissonGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
```

#### Coefficients (exponentiated)

```{r pois-coef}
mod.pois$coefficients
```

#### Model fit

```{r pois-diag, fig.height = 4}
mod.pois$diagnostics$plot
```

The dispersion ratio of `r round(mod.pois$diagnostics$dispersion_ratio, 3)`
is flagged in red — the observed variance is ~60% larger than expected under
Poisson, suggesting overdispersion. We also inspect the squared Pearson
residual plot:

```{r pois-r2, fig.height = 4, fig.width = 5}
mod.pois$diagnostics$r2_plot
```

The wedge shape confirms the mean-variance relationship is not well captured
by the Poisson assumption.

### 3. Negative Binomial Regression

The negative binomial adds a free dispersion parameter $\theta$ to handle
overdispersion.

```{r negbin}
mod.nb <- negbinGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat,
  control = stats::glm.control(maxit = 100)
)
```

#### Coefficients (exponentiated)

```{r nb-coef}
mod.nb$coefficients
```

#### Model fit

```{r nb-diag, fig.height = 4}
mod.nb$diagnostics$plot
```

```{r nb-r2, fig.height = 4, fig.width = 5}
mod.nb$diagnostics$r2_plot
```

The dispersion ratio is now
`r round(mod.nb$diagnostics$dispersion_ratio, 3)` — much closer to 1. The
estimated $\theta$ is `r round(mod.nb$theta, 3)`.

### 4. Comparing Models: Likelihood Ratio Test

Because the Poisson model is nested within the negative binomial (Poisson is
NB with $\theta \to \infty$), we can use a likelihood ratio test. The
underlying `glm`/`glm.nb` fit objects are available via `$model`:

```{r lrt}
lmtest::lrtest(mod.pois$model, mod.nb$model)
```

The negative binomial model is significantly better ($p < 0.0001$), confirming
that overdispersion is a genuine problem for the Poisson fit.

### 5. Using the `countGLM` Wrapper

Rather than fitting each model manually, `countGLM()` fits all four families
at once and selects the best by AIC — arriving at the same conclusion
automatically.

#### Overall EQI formula

```{r countglm1}
result1 <- countGLM(
  FinalEC ~ eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
print(result1)
```

The wrapper selects the same winner as the manual LRT. Individual fits remain
accessible: `result1$fits$negbin`, `result1$fits$poisson`, etc.

#### Sub-index formula

The researchers also evaluated whether separate Water, Air, Land, and
Socioeconomic indices were more informative than the composite Overall EQI.
`countGLM` handles this in one call:

```{r countglm2}
result2 <- countGLM(
  FinalEC ~ water_eqi_2jan2018_vc +
    land_eqi_2jan2018_vc +
    air_eqi_2jan2018_vc +
    sociod_eqi_2jan2018_vc +
    pctnonwhite10 +
    metro +
    gdp2017b +
    fac_penalty_count +
    CIDDist +
    EPAregion,
  data = Greenberg26.dat
)
print(result2)
```

Again the negative binomial is selected. The non-nested comparison between
these two winning models (overall EQI vs sub-indices) can be done with a Vuong
test via `nonnest2::vuongtest(result1$fits$negbin$model, result2$fits$negbin$model)`.

---

## 6. Coefficient Interpretation

`interpret_coef()` translates any exponentiated coefficient into a
plain-language statement, with a 95% CI and an automatic note when the
predictor is not discernible from zero (p > 0.05).

#### Significant predictor

```{r interp-pct}
interpret_coef(mod.nb, "pctnonwhite10")
```

#### Non-significant predictor

```{r interp-eqi}
interpret_coef(mod.nb, "eqi_2jan2018_vc")
```

The note about discernibility is added automatically.

#### Using with `countGLM`

Pass the `countGLM` result directly — it delegates to the best-fitting model:

```{r interp-countglm}
interpret_coef(result2, "pctnonwhite10")
```

#### Zero-inflated models: specifying the component

For zero-inflated models, use `component = "count"` (default) or
`component = "zero"`:

```{r interp-zi, eval = FALSE}
interpret_coef(zi_fit, "pctnonwhite10", component = "count")
interpret_coef(zi_fit, "pctnonwhite10", component = "zero")
```

---

## Case Study 2: Urban Surveillance Camera Counts (Dahir 2025)

Dahir (2025) models the number of surveillance cameras per census tract
across ten US cities. The response `cam_count` is a non-negative integer
with 87.6% zeros and a variance-to-mean ratio of approximately 3.6 —
classic signs of both excess zeros and overdispersion.

```{r load-dahir}
data("Dahir25.dat")
```

### 7. Data Exploration

```{r dahir-summarize, fig.height = 5}
summarizeCountData(
  cam_count ~ total_crime_rate + hinc + modal_zone + pvac,
  data = Dahir25.dat
)
```

The summary table makes the issue concrete: the variance-to-mean ratio far
exceeds 1 and the frequency table is dominated by zeros. Camera placement
varies substantially by land-use zone — mixed and industrial tracts have
cameras in roughly 28–42% of cases, while residential tracts (the
majority, n = 8913) have cameras in fewer than 10%. This mixture of
structurally camera-free tracts alongside genuinely high-count commercial
zones is exactly the setting where zero-inflated or overdispersed count
models are needed.

### 8. Automatic Model Selection with `countGLM`

`countGLM()` fits all four count families and selects the best by AIC. Road
length (`log_road_length`) is included as an offset in the count component;
because `ziformula = NULL`, it is automatically applied to the zero-inflation
component as well — `countGLM` prints a note confirming this:

```{r dahir-countglm}
result_cam <- countGLM(
  cam_count ~ pnhblk + pnhwht + total_crime_rate + hinc + pvac + modal_zone +
    offset(log_road_length),
  data = Dahir25.dat
)
print(result_cam)
```

The AIC and BIC tables show the ranking across all four families. The
selected model is **`r result_cam$best_model`**. The recommendation note
captures the relevant dispersion and zero-inflation diagnostics
automatically, explaining why this family was preferred.

### 9. Interpreting the Winning Model

`interpret_coef()` delegates to the best-fitting model automatically:

```{r dahir-interp-crime}
interpret_coef(result_cam, "total_crime_rate")
```

```{r dahir-interp-income}
interpret_coef(result_cam, "hinc")
```

For zero-inflated models, the count and zero components can be interpreted
separately. The count component describes the expected camera count among
tracts that are in the counting process; the zero component describes the
odds of being a structural zero (a tract that never receives a camera):

```{r dahir-interp-zi, eval = inherits(result_cam$fits[[result_cam$best_model]], "zeroinflGLMfit")}
interpret_coef(result_cam, "total_crime_rate", component = "count")
interpret_coef(result_cam, "total_crime_rate", component = "zero")
```

---

## References

Dahir, Abdi Latif. 2025. "Surveillance Camera Placement and Urban
Inequality." Working paper.

Greenberg, Pierce, Erik W Johnson, Jennifer Schwartz, and Rylie Wartinger.
2026. "Social Factors Shape Federal Environmental Crime Prosecution Patterns
in the USA." *Nature Sustainability*, 1–5.
